---
title: "Posit Academy: Week 6 Demo"
subtitle: "Quick intro to `purrr::map()`."
author: "Grace Lawley"
date: "2024-10-21"
format: html
---


```{r}
# load packages
library(purrr)
``` 


## map()

A function from the purrr package in the tidyverse

```
map(.x, .f)
```

- Inputs
    - .x = a list or vector
    - .f = a function
- Applies `.f`, to each element of `.x`.
- `map()` always returns a list



### Example 1

`.f` can be a named function, such as `rnorm()`

```{r}
map(1:5, rnorm, n = 10)
```


### Example 2

`.f` can also be an anonymous function, e.g.:

  - `\(x) x + 1`
  - `function(x) x + 1`

```{r}
map(1:5, \(x) rnorm(n = 10, x))
```

## map() variants

- Variants return a vector of the indicated type (or die trying)
- `map_lgl()`: returns a logical vector
- `map_int()`: returns a numeric vector of integers
- `map_dbl()`: returns a numeric vector of doubles
- `map_chr()`: returns a character vector


```{r}
1:5 |>
  map(rnorm, n = 10) |>  # output a list
  map(mean) # output a list
```


```{r}
1:5 |>
  map(rnorm, n = 10) |>  # output a list
  map_dbl(mean) # output a vector of doubles
```


## Resources

- This chapter of the Advanced R book: https://adv-r.hadley.nz/functionals.html#map
- purrr::map(): https://purrr.tidyverse.org/reference/map.html

