---
title: "Week 5: Work with data types"
author: "Andrew Heiss"
date: "2026-02-24"
---

## Recoding variables

There are lots of ways to recode variables! Here are some:

```{r}
#| warning: false
#| message: false

library(tidyverse)
```

### `ifelse()` (don't do this!)

```{r}
mpg_cleaned_ifelse <- mpg |>
  mutate(
    class_clean = ifelse(
      class == "compact",
      "Compact",
      ifelse(
        class == "midsize",
        "Midsize",
        ifelse(
          class == "suv",
          "SUV",
          "something else"  #  AND SO ON for all the other classes
        )
      )
    )
  )

mpg_cleaned_ifelse |> 
  select(class, class_clean, everything())
```

### `case_when()` (do this!)

`case_when()` is like a fancier, multi-statement version of `ifelse()`:

```{r}
mpg_cleaned_case_when <- mpg |>
  mutate(
    class_clean = case_when(
      class == "compact" ~ "Compact",
      class == "midsize" ~ "Midsize",
      class == "suv" ~ "SUV",
      .default = "Something else"  # Anything that doesn't match gets this value
    )
  )

mpg_cleaned_case_when |> 
  select(class, class_clean, everything())
```

### `case_match()` (do this!)

`case_match()` is like `case_when()`, but for only running tests in one variable:

```{r}
mpg_cleaned_case_match <- mpg |>
  mutate(
    class_clean = case_match(
      class,
      "compact" ~ "Compact",
      "midsize" ~ "Midsize",
      "suv" ~ "SUV",
      .default = "Something else"
    )
  )

mpg_cleaned_case_match |>
  select(class, class_clean, everything())
```

### `left_join()` another dataset (do this!)

You can also use a smaller dataset and left join it into the main dataset. This is helpful when you have lots of things to recode (like patient-level information):

```{r}
# Make a little lookup dataset mapping class to nicer names
classes <- tribble(
  ~class       , ~class_clean ,
  "compact"    , "Compact"    ,
  "midsize"    , "Midsize"    ,
  "suv"        , "SUV"        ,
  "2seater"    , "Two-seater" ,
  "minivan"    , "Minivan"    ,
  "pickup"     , "Pickup"     ,
  "subcompcat" , "Subcompact"
)

# Join it
mpg_cleaned_left_join <- mpg |>
  left_join(classes, by = join_by(class))

mpg_cleaned_left_join |>
  select(class, class_clean, everything())
```

### `recode()`

`recode()` works too, but the documentation for it recommends using `case_when()` or `case_match()`:

```{r}
mpg |>
  mutate(
    class_clean = recode(class, "compact" = "Compact", "midsize" = "Midsize")
  )
```

## Milestone start

```{.r}
treatment <- mdrd_tidy %>% 
  separate_wider_delim(cols = treatment, delim = "_", names = c("diet", "bp")) %>% 
  # You could clean up the columns first and remove the word "diet", but you 
  # could also just leave "diet" in there and match in case_match() based on 
  # the full original value
  # mutate(diet = str_remove(diet, "diet"))
  mutate(diet_clean = case_match(
    diet,
    "dietl" ~ "Low-protein",
    "dietk" ~ "Very low-protein"
  ))
```
