---
title: "Posit Academy: Week 2 Demo"
subtitle: "Two ways to filter"
author: "Grace Lawley"
date: "2024-09-15"
format: html
---
  
```{r}
# load packages
library(tidyverse)

# read in covid dataset
covid <- read_csv("../data/covid.csv")
```


## Question 1

> Q: If we want to filter `covid` so we only have data for specific states, what's the easiest way to do this?

A couple of options are shown below. I've included `count(state)` to check that the resulting filtered dataframe is what we are expecting.

### Option 1: combine multiple logical tests using & or |

```{r}
covid |> 
  filter(state == "OR" | state == "WA" | state == "CA") |> 
  count(state)
```

Pros/Cons:

-   Clear, easy to read
-   Repetitive, lots of typing
-   Not ideal if filtering for many states

### Option 2: use the built-in %in% operator

> Note: to look up the help documentation for the %in% operator, type the following into the console: ?`%in%`

```{r}
covid |>
  filter(state %in% c("OR", "WA", "CA")) |>   # filter for 3 states
  count(state)                                # get counts for each unique state
```

Pros/Cons:

-   Ideal if filtering for many states
-   Can create vector with states to include *outside* of pipe

```{r}
states_to_include <- c("OR", "WA", "CA")

covid |> 
  filter(state %in% states_to_include) |> 
  count(state)
```




