---
title: "Posit Academy: Week 3 Demo"
subtitle: "Introduction to ggplot2"
author: "Grace Lawley"
date: "2024-09-22"
format: html
---
  
```{r}
# load packages
library(tidyverse)

# read in covid dataset
covid <- read_csv("../data/covid.csv")

# only data from OR, WA, and CA
covid_west <- covid %>%
  filter(state %in% c("OR", "WA", "CA"))
```


## `geom_line()`

To create a minimal line plot:

```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line()
```

## Themes

ggplot2 has lots of themes included that work out of the box, such as `theme_bw()`, `theme_linedraw()`, and `theme_minimal()`.

You can read more about theme here: https://ggplot2.tidyverse.org/reference/ggtheme.html

```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line(alpha = 0.8) + # alpha changes the transparency of the line
  theme_minimal()
```


## `facet_wrap()`

You can "facet" a plot by a variable in your dataset by using `facet_wrap`. 
The `ncol` and `nrow` arguments let you adjust how many columns or rows you want to have.

```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line() +
  facet_wrap(vars(state), ncol = 1)
```
```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line() +
  facet_wrap(vars(state), nrow = 2)
```

When faceting, both axes are fixed by default (i.e., the same axis is shared across all plots).
You can change this by setting the `scales` to `"free"`, `"free_x"`, or `"free_y"`.

```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line() + 
  facet_wrap(vars(state), ncol = 1, scales = "free") 
```

## `labs()`

You can use the `labs()` function to rename the axes, add a title, subtitle, caption, etc.

```{r}
ggplot(covid_west, aes(x = date, y = cases, color = state)) +
  geom_line() + 
  facet_wrap(vars(state), ncol = 1) +
  labs(title = "COVID Cases in West Coast States",
       x = NULL,
       caption = "Source: https://covidtracking.com/",
       color = NULL)
```
