---
title: "Control function flow"
subtitle: "Week 2"
author: "Andrew Heiss"
date: "2025-05-20"
format:
  html:
    toc: true
---

## Vectors and lists

```{r}
# Different ways to make vectors
x <- c(1, 2, 3, 4, 5)
x <- 1:5
x <- seq(1, 5, by = 1)
x <- seq(1, 5, by = 0.1)

# All items in a vector have to be the same type
# Adding a character "3" forces R to make all the other elements characters
x <- c(1, 2, "3", 4, 5)
x

# ... which means you can do math with it anymore
mean(x)

# You can name slots in a vector if you want
x <- c(first = 1, second = 2, red = 3, 4, 5)
x

# And you can access elements by name
x["second"]


# Lists are collections of things, and list elements can be any type and any length
z <- list(1:5, "A")
z

# List elements can be named and you can reference them by name
z <- list(numbers = 1:5, letter = "A")
z$numbers
z$letter


# Most R objects are secretly just lists with lots of different slots
model <- lm(mpg ~ cyl, data = mtcars)

model$coefficients
model$residuals

# You can access deeply nested list elements too, by name...
model$qr$qr
# ...or by position...
model[[7]][[1]]
```

## List columns

Data frames can contain list columns, which is super fun and cool because you can keep deeply nested lists all together in one place and iterate through them.

Here is a data frame with three rows, with a column named "data" that contains a complete subset of the `mtcars` data:

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(broom)  # For extracting stuff from model objects

qwer <- mtcars |> 
  group_by(cyl) |> 
  nest()
qwer
```

We can then run `lm()` on each of the smaller datasets and extract pieces of that model object. The "model" column here contains the three regression models, and "coef_thing" contains the coefficient for `disp`

```{r}
qwer <- mtcars |> 
  group_by(cyl) |> 
  nest() |> 
  mutate(model = map(data, \(x) lm(mpg ~ disp, data = x))) |> 
  mutate(coef_thing = map_dbl(model, ~ .$coefficients[[2]]))
qwer
```

↑ That's the same as doing this, but these are all separate objects that you now have to keep track of:

```{r}
thing1 <- lm(mpg ~ disp, data = filter(mtcars, cyl == 6))
thing1$coefficients[[2]]
thing2 <- lm(mpg ~ disp, data = filter(mtcars, cyl == 4))
thing2$coefficients[[2]]
thing3 <- lm(mpg ~ disp, data = filter(mtcars, cyl == 8))
thing3$coefficients[[2]]
```

You can use `unnest()` to convert a list column back to a regular long data frame:

```{r}
qwer <- mtcars |> 
  group_by(cyl) |> 
  nest() |> 
  mutate(model = map(data, \(x) lm(mpg ~ disp, data = x))) |> 
  mutate(coef_thing = map(model, ~ tidy(., conf.int = TRUE))) |> 
  unnest(coef_thing)
qwer
```

And then you can filter it and mutate it and plot it and do whatever other normal data manipulation stuff you want:

```{r}
qwer |> 
  filter(term == "disp") |> 
  ggplot(aes(x = estimate, y = factor(cyl))) + 
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(title = "Coefficient for disp across cylinders")
```

## Milestone start

```{{r}}
data |>
  select(division, where(is.numeric), -year) |>
  pivot_longer(-division) |>
  filter(name == var) |>
  group_by(division, name) |>
  summarize(value = sum(value, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(value)) |>
  pivot_wider()
```
