---
title: "Week 3: Write functions with the tidyverse"
author: "Andrew Heiss"
date: "2025-09-23"
---

```{r}
#| include: false
options(width = 300)
```

## List columns

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(broom)

penguins <- penguins |> drop_na(sex)
```

List columns can contain any kind of variable, like whole data frames. Here, the `data` column contains three datasets: all the rows of the `penguins` data for each of the three species.

```{r}
penguins |> 
  group_by(species) |> 
  nest()
```

We can feed that list column to `map()` and make new columns. Like here's the same linear regression model fit to each of the three datasets:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x)))
```

We can work with that list of `lm` objects too, like feeding them through `tidy()` to create a data frame of model results:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE)))
```

If we unnest that `results` column we'll get a long data frame with one row for each of the rows in the data:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  unnest(results)
```

## Special `map()` variants

We can also extract values from the nested `results` data frames, like finding the flipper length coefficients for each species:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  mutate(
    flipper_coef = map(
      results,
      \(x) x |> filter(term == "flipper_length_mm") |> pull(estimate)
    )
  )
```

This new `flipper_coef` column contains single numbers, but they're weirdly wrapped up inside a list. Since each list element only contains one item, and that one item is a number, we can use `map_dbl()` instead of `map()` to tell R that the output really is just a number:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  mutate(
    flipper_coef = map_dbl(
      results,
      \(x) x |> filter(term == "flipper_length_mm") |> pull(estimate)
    )
  )
```

Now that `flipper_coef` column is an actual number.

These special `map()` variants only work when the results of the mapped function are of length 1. For instance, here we'll pull out both values from the `estimate` column by not filtering it beforehand:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  mutate(
    coefs = map(
      results,
      \(x) x |> pull(estimate)
    )
  )
```

That returns a vector of two numbers for each row (or `<dbl [2]>`). If we use `map_dbl()`, we'll get an error:

```{r}
#| error: true
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  mutate(
    coefs = map_dbl(
      results,
      \(x) x |> pull(estimate)
    )
  )
```

We can use `map_chr()` and `map_lgl()` to do the same thing for character and logical (TRUE/FALSE) values:

```{r}
penguins |>
  group_by(species) |>
  nest() |>
  mutate(
    model = map(data, \(x) lm(body_mass_g ~ flipper_length_mm, data = x))
  ) |>
  mutate(results = map(model, \(x) tidy(x, conf.int = TRUE))) |>
  mutate(
    flipper_coef = map_dbl(
      results,
      \(x) x |> filter(term == "flipper_length_mm") |> pull(estimate)
    ),
    coef_name = map_chr(
      results,
      \(x) x |> filter(term == "flipper_length_mm") |> pull(term)
    ),
    has_two_rows = map_lgl(results, \(x) nrow(x) == 2)
  )
```

::: {.callout-tip}

### Use `map()` first!

I always mess up when using `map_*()` the first time—like, I'll try it when a function returns a vector or a data frame or something else instead of a single value. 

My strategy is to always use `map()` first. If I then see that it's returning a nested list of one-element things (like `<dbl [1]>` or `<int [1]>` or `<chr [1]>` or `<lgl [1]>`), then I'll change it to the corresponding special map (like `map_dbl()`, `map_int()`, `map_chr()`, or `map_lgl()`).

:::


## Milestone start + embracing and variable substitution

In the milestone, you'll create a function that takes unquoted variable names. You'll need to protect those using the `{{}}` ("embrace") [tidy eval operator](https://ggplot2.tidyverse.org/reference/tidyeval.html). This replaces the value of `.var` inside `pull()` and `ggplot()` and other functions:

```{r}
plot_var_by <- function(.data, .var = hwy, .by = class) {
  column_to_plot <- .data |> 
    pull({{.var}})

  if (is.numeric(column_to_plot)) {
    .data |>
      mutate({{.by}} := fct_reorder({{.by}}, {{.var}})) |>
      ggplot(aes(x = {{.var}}, y = {{.by}})) +
      geom_boxplot()
  } else {
    # DO OTHER STUFF
  }
}

plot_var_by(mpg, .var = hwy)

# Or with pipes
mpg |> 
  plot_var_by(hwy)
```

The base R equivalent of this *feels* like this, but **this does not work**. The `{{}}` operator only works inside tidy-enabled functions like all the {dplyr}, {tidyr}, and {ggplot2} functions.

```{r}
#| eval: false
plot_var_by <- function(.data, .var = hwy, .by = class) {
  is.numeric(.data${{.var}})
}
```

If you really want, you can do this unquoted variable substitution process with base R's `substitute()` and `eval()`:

```{r}
plot_var_by <- function(.data, .var) {
  # Base R's way of doing the embracing thing
  #
  # substitute() takes the value of the function argument `var` and replaces it 
  # with whatever is passed, and then `eval()` evaluates it and extracts it 
  # from `.data`
  column_to_plot <- eval(substitute(.var), .data)

  is.numeric(column_to_plot)
}

plot_var_by(mpg, hwy)

mpg |> 
  plot_var_by(hwy)
```

That `eval(substitute(...))` stuff is gross and complex, hence the invention of `{{}}`

## Walruses

To use an embraced variable name as the name for a variable in `mutate()` or `summarize()`, you need to use the walrus operator, or `:=`. That's because, according to the [tidy eval documentation](https://ggplot2.tidyverse.org/reference/tidyeval.html):

> For technical reasons, the R language doesn't support complex expressions on the left of `=`, so we use `:=` as a workaround.

```{r}
find_median <- function(.data, .var = hwy) {
  .data |> 
    summarize(
      {{.var}} := median({{.var}}),
      # You can use {{}} things in quotes too:
      "{{.var}}_median" := median({{.var}})
    )
}

mpg |> 
  find_median(cty)
```
